pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-19 manuscript/greenwell.tex
minor edits

The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}. According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct. For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs.

@@

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing partial-residual and marginal-model plots. The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest. However, these methods are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are mainly for nonparametric and black box models (though they can be used for simple parametric models as well).

@@

Currently, only two functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument. If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function. The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$. If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}. The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options. Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.

>>>>

The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}. According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time. They are also less accurate than PDPs, but are faster to construct. For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs.

@@

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing partial-residual and marginal-model plots. The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest; it provides tabular and graphical displays for the terms in parametric models. However, these methods were designed for and are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).

@@

Currently, only two functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument. If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function. The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$. If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}. The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options (e.g., the option to add a LOESS smooth). Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
